%══════════════════════════════════════════════════════════════════════
%  CHAPTER 1 — Introduction  (unnumbered per template)
%══════════════════════════════════════════════════════════════════════
\chap{Introduction}
\label{ch:intro}

\pagestyle{fancy}

%%─────────────────────────────────────────────────────────────────────
\sect{Motivation}
%%─────────────────────────────────────────────────────────────────────

Distributed Denial-of-Service (DDoS) attacks continue to grow in both
volume and sophistication. Industry reports record multi-terabit-per-second
volumetric floods alongside a rising tide of application-layer and
low-rate attacks that evade simple rate thresholds. In modern cloud
data-center networks (DCNs), such attacks threaten not only a single
victim tenant but also every co-tenant sharing the physical substrate:
a volumetric surge against one tenant can saturate aggregation and core
links in the fat-tree topology, triggering service-level agreement (SLA)
violations across the entire multi-tenant environment.

The dominant defensive posture today is \emph{reactive}: a detection rule
fires only after an offending feature---typically packet rate or SYN
ratio---crosses a static threshold, and only then does the control plane
instantiate a mitigation Virtual Network Function (VNF). The end-to-end
latency of this response loop is dominated by VNF boot time. Measurements
on the PAD-ONAP prototype confirm this floor at approximately 6{,}006~ms
for a full traffic scrubber; during that window the attack proceeds
unchecked. This latency is not a deficiency of any particular
orchestrator---it is \emph{structural}. Lowering it requires acting
\emph{before} the attack crosses the reactive trigger, which in turn
requires a forecaster.

At the same time, purely threshold-based detection is brittle in the face
of out-of-distribution (OOD) traffic. An HTTP flood or an ICMP burst whose
feature vector falls in an unexpected region of the threshold space may
cause a rule-based system to over-escalate---instantiating an expensive
scrubber VNF unnecessarily and degrading benign tenant throughput. An
AI-trained classifier, by contrast, can assign calibrated probabilities
and degrade gracefully on unseen patterns. These two motivating gaps---the
reactive latency floor and the OOD brittleness of threshold rules---frame
the research problem addressed in this thesis.

%%─────────────────────────────────────────────────────────────────────
\sect{Research Challenges and Research Problem}
%%─────────────────────────────────────────────────────────────────────

We consider the orchestrator of a multi-tenant DCN whose east--west and
north--south traffic is monitored via gNMI and NetFlow, whose mitigation
capacity is expressed as instantiable VNFs under an ONAP MANO control plane,
and whose tenants hold differentiated SLAs (e.g.~URLLC with a guaranteed
bandwidth floor, eMBB best-effort, mMTC elastic). The defender must decide,
for every 5-second telemetry window, \emph{which} tier of mitigation to
apply---ranging from no action to heavy scrubbing---so as to minimise attack
impact while preserving benign-tenant SLAs and keeping false-positive
actuation bounded.

Three concrete challenges arise:

\begin{enumerate}
  \item \textbf{Reactive latency floor.} The minimum time to respond equals
        the VNF boot time ($\approx 6$~s for a scrubber). No amount of
        orchestration tuning can reduce this; only pre-positioning the VNF
        \emph{before} the attack peaks can.

  \item \textbf{OOD over-escalation.} Threshold rules with fixed coefficients
        fire on any feature pattern that crosses the threshold, regardless of
        whether the pattern belongs to a known attack class. This causes
        unnecessary VNF actuation on novel traffic.

  \item \textbf{SLA collision under mitigation.} Instantiating a VNF consumes
        link capacity, which can undercut a co-tenant's SLA floor if bandwidth
        is not re-allocated explicitly.
\end{enumerate}

Existing systems address at most two of these challenges simultaneously.
Systems that detect but do not orchestrate leave challenge~1 open; purely
reactive orchestrators leave challenge~1 open; forecasters that do not
integrate with a standards-compliant MANO stack cannot enforce challenge~3.
No published system combines multi-horizon forecasting, ONAP-compliant VNF
lifecycle management, and an LP-guaranteed multi-tenant SLA allocation in
a single closed-loop pipeline.

We formalise the scope through three research questions and three testable
hypotheses.

\medskip
\noindent\textit{Research Questions:}

\begin{itemize}
  \item \textbf{RQ1 (Lead-time).} Does an ML forecaster reduce end-to-end
        mitigation latency compared with a rule-based threshold baseline at
        a controlled false-positive rate?
  \item \textbf{RQ2 (SLA preservation).} Does a graduated five-tier policy
        engine preserve per-tenant SLAs under active mitigation?
  \item \textbf{RQ3 (Safety).} Is proactive pre-positioning at \tier{2}
        safe---does it not degrade benign tenants below their SLA floor,
        and can it be rolled back within a bounded number of windows?
\end{itemize}

\medskip
\noindent\textit{Hypotheses:}

\begin{itemize}
  \item \textbf{H1.} On gradual-ramp attacks (SYN ramp), the ML forecaster
        triggers mitigation at least 30 seconds earlier than a rule-based
        threshold baseline.
  \item \textbf{H2.} Binary detection accuracy $\geq 0.98$ and macro AUC
        $\geq 0.99$ are achievable under a strict temporal 80/20 train--test
        split with no data shuffling.
  \item \textbf{H3.} Under all eight evaluation scenarios the URLLC bandwidth
        floor of 150~Mbps is preserved at every mitigation tier.
\end{itemize}

%%─────────────────────────────────────────────────────────────────────
\sect{Objective and Contents}
%%─────────────────────────────────────────────────────────────────────

The objective of this thesis is to design, implement, and empirically evaluate
\padonap{} (\textit{Proactive Attack Defense over ONAP}), an AI-augmented
NFV orchestration pipeline that satisfies the following requirements:

\begin{itemize}
  \item \textbf{Proactive response.} The system shall pre-position
        mitigation VNFs based on a multi-horizon ML forecast before the
        attack reaches the detection threshold.
  \item \textbf{ONAP compliance.} All VNF lifecycle operations shall be
        issued through the ONAP SO v7 REST API; the AI and orchestration
        layers shall be decoupled via a canonical JSON schema on the DMaaP
        message bus.
  \item \textbf{SLA guarantee.} A linear-programming allocator shall
        guarantee per-tenant bandwidth floors under any active mitigation tier.
  \item \textbf{OOD robustness.} The system shall degrade gracefully
        (under-respond rather than over-escalate) on traffic distributions
        absent from the training set.
\end{itemize}

%%─────────────────────────────────────────────────────────────────────
\sect{Contribution Highlights}
%%─────────────────────────────────────────────────────────────────────

This thesis makes four technical contributions:

\begin{enumerate}
  \item \textbf{A four-stage AI-augmented architecture} (\texttt{pipeline/})
        comprising a gNMI/NetFlow telemetry collector (S1), a 17-feature
        entropy and rate extractor (S2), a dual-track AI inference layer
        (S3), and a five-tier graduated orchestration layer with ONAP
        lifecycle management (S4). A canonical \texttt{AIOutputPayload}
        JSON schema published on DMaaP decouples the AI and orchestration
        layers, enabling independent evolution of each component.

  \item \textbf{A Transformer+LSTM multi-horizon DDoS forecaster} producing
        attack-probability estimates at $t{+}30, 60, 90, 120$ seconds. The
        AUC degrades monotonically from 0.988 at $t{+}30$~s to 0.971 at
        $t{+}120$~s, consistent with the information-decay expected in
        forecasting and indicative of a well-calibrated model.

  \item \textbf{A linear-programming multi-tenant SLA allocator}
        (\texttt{sla\_allocator.py}) that, at every tier transition, solves a
        bounded LP to guarantee per-tenant bandwidth floors. The URLLC floor
        of 150~Mbps is preserved in all eight evaluation scenarios.

  \item \textbf{A quantitative comparison against a rule-based baseline}
        showing an 8/8 vs 6/8 scenario pass rate, a 65-second lead-time
        advantage on gradual-ramp attacks, and a $10.9\times$ proactive
        latency speed-up (505~ms vs 6{,}006~ms).
\end{enumerate}

%%─────────────────────────────────────────────────────────────────────
\sect{Thesis Outline}
%%─────────────────────────────────────────────────────────────────────

The remainder of the thesis is organised as follows.

Chapter~\ref{ch:background} surveys the technical foundations: DDoS attack
taxonomy in DCNs, NFV and Service Function Chaining, the ONAP MANO
architecture, machine-learning models for network security, forecasting
methodologies, and fat-tree DCN topology.

Chapter~\ref{ch:related} reviews related work, positions \padonap{} against
eight recent systems on a six-axis comparison matrix, and identifies the
three research gaps that motivate this work.

Chapter~\ref{ch:system} presents the \padonap{} architecture in full detail:
system overview, formal threat model, safety invariants, five-tier policy
state machine, LP SLA allocator formulation, and the AIOutputPayload
interface contract.

Chapter~\ref{ch:ai} details the AI components: the 17-feature vector, the
XGBoost 7-class detector (Acc = 0.9825, Macro-AUC = 0.9958), the
Transformer+LSTM 4-horizon forecaster, three cross-validation strategies,
FGSM adversarial augmentation, and SHAP interpretability.

Chapter~\ref{ch:eval} reports the full experimental evaluation: the eight
orchestration scenarios, the rule-based threshold baseline comparison,
lead-time analysis, SLA fairness verification, and a state-of-the-art
comparison.

The Conclusion summarises the contributions, confirms the three hypotheses,
and outlines five directions for future work.
