%══════════════════════════════════════════════════════════════════════
%  CHAPTER 2 — Background
%══════════════════════════════════════════════════════════════════════
\chapter{Background}
\label{ch:background}

This chapter introduces the technical foundations required to understand
and contextualise \padonap{}. Sections~\ref{sec:ddos-bg}--\ref{sec:fttopo}
cover DDoS attack taxonomy, Network Function Virtualization and Service
Function Chaining, the ONAP MANO ecosystem, machine-learning approaches
for network security, multi-horizon forecasting, and fat-tree data-center
topology. Section~\ref{sec:bg-summary} summarises the key concepts.

%%─────────────────────────────────────────────────────────────────────
\section{DDoS Attacks in Data-Center Networks}
\label{sec:ddos-bg}

A Distributed Denial-of-Service (DDoS) attack directs traffic from a
large number of compromised hosts (a \emph{botnet}) at a victim,
consuming link bandwidth, server CPU cycles, or protocol state until
legitimate requests are dropped. In the data-center context three
families are most relevant.

\textbf{Volumetric floods} --- UDP floods, ICMP floods, and amplification
attacks (DNS, NTP, memcached reflection) --- aim to exhaust link capacity
at the DCN edge or aggregation layer. Attack rates routinely exceed
1~Tbps, far beyond the capacity of a single victim server.

\textbf{Protocol-exhaustion attacks} --- SYN floods and TCP-state
exhaustion --- target finite state machines in middleboxes and server
OS kernels. A SYN flood, for instance, fills the TCP half-open connection
table by sending millions of SYN packets and never completing the
three-way handshake.

\textbf{Application-layer attacks} --- HTTP floods and Slow-rate attacks ---
mimic legitimate traffic patterns at high or low rate to evade network-layer
filtering. These attacks are notoriously difficult to distinguish from
flash-crowd events and require semantic analysis of request content or
timing behaviour.

CICDDoS2019~\citep{sharafaldin2019cicddos} is the most widely cited
public dataset covering the first two families. Application-layer samples
carry label metadata but have incomplete flow records, which constrains
empirical evaluations and motivates the multi-class design described in
Chapter~\ref{ch:ai}.

%%─────────────────────────────────────────────────────────────────────
\section{Network Function Virtualization and Service Function Chaining}
\label{sec:nfv-bg}

Network Function Virtualization (NFV), standardised by ETSI NFV ISG
\citep{etsi_nfv_arch}, decouples network packet-processing functions from
specialised proprietary hardware. A \emph{Virtual Network Function} (VNF)
is a software implementation of a network function (e.g.~a firewall,
rate-limiter, or traffic scrubber) that runs on commodity servers under a
hypervisor or container engine. NFV enables elastic scaling: a VNF instance
can be instantiated on demand and terminated when no longer needed.

Service Function Chaining (SFC), specified in IETF RFC~7665, composes
ordered sequences of VNFs into \emph{chains} through which traffic is
steered. OpenFlow and its Open vSwitch (OVS) implementation provide the
southbound mechanism to insert and remove SFC steering rules in a
programmable data plane. For DDoS mitigation the chain typically consists
of a \emph{rate-limiter} (low overhead, fast boot) at the head and a
\emph{scrubber} (deep packet inspection, slower boot) at the tail.

A key design parameter is \emph{VNF boot latency}: the time from issuing
an instantiation request to the VNF being ready to handle traffic. For a
lightweight rate-limiter this is approximately 500~ms; for a full scrubber
it is approximately 6~s, a fundamental difference that motivates the
five-tier design of \padonap{}.

%%─────────────────────────────────────────────────────────────────────
\section{ONAP MANO Architecture}
\label{sec:onap-bg}

The Open Network Automation Platform (ONAP) is the Linux Foundation
reference orchestrator for NFV management and network automation. ONAP
implements the ETSI NFV MANO framework with the following sub-components
relevant to DDoS closed-loop response:

\begin{itemize}
  \item \textbf{Service Orchestrator (SO)} instantiates and terminates VNF
        instances through a REST API
        (\texttt{/onap/so/infra/serviceInstantiation/v7/}).
        The v7 API supports synchronous and asynchronous lifecycle
        operations; \padonap{} uses synchronous calls for latency
        measurement.

  \item \textbf{CLAMP} (Closed Loop Automation Management Platform)
        manages closed-loop templates and pushes operational policies to
        the Policy Framework. In \padonap{}, CLAMP subscribes to the DMaaP
        AI topic and translates \texttt{AIOutputPayload} events into SO
        instantiation calls.

  \item \textbf{Policy Framework} evaluates Drools rules and issues
        operational policies that CLAMP and SO consume. The five-tier
        policy engine of \padonap{} is modelled on the Policy Framework
        evaluation logic.

  \item \textbf{DMaaP} (Data Management as a Platform) is ONAP's
        Kafka-based event bus. AI decisions are published as JSON events
        on topic \texttt{PAD\_ONAP\_AI\_SIGNALS}; downstream ONAP
        components subscribe and react without tight coupling to the AI
        layer.
\end{itemize}

The key architectural value of ONAP for \padonap{} is the \emph{closed
loop}: telemetry events flow from the network into CLAMP, trigger policy
evaluation, drive SO instantiation, and the resulting VNF behaviour
changes the telemetry observed in the next window. This loop is what
enables proactive pre-positioning: the AI can influence VNF state before
the attack is fully detected.

%%─────────────────────────────────────────────────────────────────────
\section{Machine Learning for Network Security}
\label{sec:ml-bg}

Three model families dominate the literature on ML-based DDoS detection.

\textbf{Tree ensembles} (XGBoost~\citep{chen2016xgboost}, LightGBM,
Random Forests) offer strong accuracy on tabular flow features, native
support for mixed-type inputs, and sub-millisecond batch-1 inference
latency. They are the current state of the art on CICDDoS2019 and
similar flow-feature datasets. XGBoost uses gradient boosting over
classification trees and supports multi-class softmax objectives
directly.

\textbf{Deep sequence models} (LSTM, Transformer~\citep{vaswani2017attention})
capture temporal structure across successive windows and naturally extend
to multi-horizon forecasting. The Transformer's self-attention mechanism
learns long-range dependencies without the vanishing-gradient limitation
of vanilla RNNs; stacking a Transformer encoder with an LSTM combines
global context (attention) with local memory (recurrence).

\textbf{Graph neural networks} model host-to-host traffic interaction as a
dynamic graph but require flow-level graph construction per window, which
is computationally expensive at DCN scale. We do not use GNNs in \padonap{}
but acknowledge them as a future direction.

\textbf{Adversarial robustness.} Goodfellow \textit{et al.}~\citep{goodfellow2015fgsm}
introduced the Fast Gradient Sign Method (FGSM) for generating adversarial
examples: $\tilde{\mathbf{x}} = \mathbf{x} + \varepsilon \cdot
\text{sign}(\nabla_{\mathbf{x}} \mathcal{L})$. Training on FGSM-perturbed
inputs improves model robustness to small feature perturbations, a
relevant concern when an adversary can slightly manipulate inter-arrival
times or packet sizes to evade detection.

\textbf{Interpretability.} SHAP (SHapley Additive exPlanations)
\citep{lundberg2017shap} provides feature attribution values grounded in
cooperative game theory. For each prediction, SHAP assigns a scalar
credit to each feature proportional to its marginal contribution. This
is critical for operator trust: a DDoS detector that identifies
\texttt{dst\_port\_entropy} as the dominant feature can be validated
against the known behaviour of DDoS floods (single-port concentration).

%%─────────────────────────────────────────────────────────────────────
\section{Multi-Horizon Forecasting}
\label{sec:forecast-bg}

A \emph{forecaster} predicts attack probability at a future time
$t + h$ given observations up to time $t$. The predictable lead-time is
bounded above by the forecast horizon $h$ and below by the VNF boot
latency $\ell_{\text{boot}}$. For the system to gain a proactive
advantage it is necessary that $h > \ell_{\text{boot}}$; choosing $h$
too large, however, degrades AUC because the information content of
current telemetry decays with horizon.

Formally, let $\{f_1, f_2, \ldots, f_n\}$ be a sequence of
17-dimensional feature vectors sampled at 5-second intervals. A
four-horizon forecaster outputs a vector
\[
  (p_{t+30},\; p_{t+60},\; p_{t+90},\; p_{t+120}) \in [0,1]^4,
\]
where $p_{t+h}$ is the predicted probability of an attack at time
$t + h$. Labels are constructed from truly future windows---not
same-window labels---to avoid circular data leakage.

A well-calibrated forecaster exhibits \emph{monotonic AUC degradation}:
$\text{AUC}(t{+}30) > \text{AUC}(t{+}60) > \text{AUC}(t{+}90) >
\text{AUC}(t{+}120)$. Non-monotonic AUC is a red flag for data leakage,
as it implies that longer-horizon labels carry more information than
short-horizon ones---physically impossible without look-ahead.

%%─────────────────────────────────────────────────────────────────────
\section{Fat-Tree Data-Center Topology}
\label{sec:fttopo}

The fat-tree topology introduced by Al-Fares \textit{et al.} is the
dominant DCN design for large-scale cloud deployments. A fat-tree with
parameter $k$ consists of $(5/4)k^2$ switches organised into three
layers: \emph{edge} switches connect servers, \emph{aggregation} switches
interconnect edge pods, and \emph{core} switches interconnect all
aggregation pods. The topology supports up to $(k/2)^2$ distinct equal-cost
paths between any server pair, enabling load balancing and fault tolerance.

In the context of DDoS mitigation, a volumetric attack targeting a server
in pod $p$ can saturate the aggregation links within pod $p$ and spill
onto core links, affecting servers in all other pods. SFC rules must
therefore be installed at the correct point in the tree to intercept
attack traffic before it reaches the aggregation or core layer.

\padonap{} is evaluated on a Mininet fat-tree with $k{=}4$ (20 switches,
16 hosts), following the classical Al-Fares design. This allows
measurement of SFC rule installation latency at 1, 2, and 3 pod levels---a
measurement absent from baselines that assume single-switch deployment.

%%─────────────────────────────────────────────────────────────────────
\section{Chapter Summary}
\label{sec:bg-summary}

This chapter established the key background concepts underpinning \padonap{}:
(i) three DDoS attack families and their distinguishing traffic signatures;
(ii) NFV/SFC as the mechanism for elastic VNF deployment, with a critical
dependence on boot latency;
(iii) ONAP's closed-loop architecture as the standards-compliant orchestration
layer;
(iv) XGBoost and Transformer+LSTM as the ML families best suited to
tabular detection and temporal forecasting respectively;
(v) the multi-horizon forecasting principle and its proactive-advantage
condition $h > \ell_{\text{boot}}$;
(vi) fat-tree topology and its implications for SFC path selection.
These concepts are built upon directly in the proposed system design
(Chapter~\ref{ch:system}) and AI component design (Chapter~\ref{ch:ai}).
